\subsection{Shared objects access}

\begin{frame}[t,fragile]{Data races}
\begin{itemize}
  \item Two potentially concurrent accesses to a memory location constitute
        a \textbad{data race} if at least one of them is a write and 
        an ordering cannot be guaranteed.
    \begin{itemize}
      \item A program with a data race has \textbad{undefined behavior}.
    \end{itemize}
\pause
\begin{lstlisting}
int x = 0;
void f() {
  ++x;
}

void g() {
  using namespace std;
  thread t1{f}; thread t2{f};
  t1.join(); t2.join();
  cout << x << "\n";
}
\end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[t,fragile]{Mutual exclusion}
\mode<presentation>{\vspace{-.75em}}
\begin{itemize}
  \item \cppid{mutex} provides access control to a resource with mutual exclusion.
    \begin{itemize}
      \item \cppid{lock()}: Acquires the associated lock.
      \item \cppid{unlock()}: Releases the associated lock.
    \end{itemize}
\begin{lstlisting}
std::mutex m;
int x = 0;

void f() {
  m.lock();
  ++x;
  m.unlock();
}

void g() {
  using namespace std;
  thread t1{f}; thread t2{f};
  t1.join(); t2.join();
  cout << x << endl;
}
\end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[t,fragile]{Problems with \texttt{lock()}/\texttt{unlock()}}
\mode<presentation>{\vspace{-.75em}}
\begin{itemize}
  \item Possible problems:
    \begin{itemize}
      \item Forgetting to release a lock.
      \item Exceptions.
    \end{itemize}
  \item Solution: \cppid{lock\_guard}.
\begin{lstlisting}
std::mutex m;
int x = 0;

void f() {
  std::lock_guard<std::mutex> l{m}; // Acquires the lock
  ++x;
} // Releases the lock

void g() {
  using namespace std;
  thread t1(f); thread t2(f);
  t1.join(); t2.join();
  cout << x << endl;
}
\end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[t,fragile]{Acquiring multiple \texttt{mutex}es}
\mode<presentation>{\vspace{-0.5em}}
\begin{itemize}
  \item Variadic \cppid{lock()} function allows acquiring
        several \cppid{mutex} at once.
    \begin{itemize}
      \item All or nothing acquisition.
      \item If any of them cannot be acquired waits releasing all of them.
    \end{itemize}
\begin{lstlisting}
std::mutex m1, m2, m3;
void f() {
  std::lock(m1, m2, m3);
  // Shared object access

} // Care: mutexes are not unlocked
\end{lstlisting}

  \pause
  \item Specially useful in cooperation with \cppid{unique\_lock}
\begin{lstlisting}
void f() {
  using namespace std;
  unique_lock<mutex> l1{m1, defer_lock}, l2{m2, defer_lock};
  lock(l1, l2);
  // Shared objects access

} // Mutexes are unlocked
\end{lstlisting}
\end{itemize}
\end{frame}

